---
title: Human protein biomarkers, with clinical relevance inferred from Oracle HealthFacts
  via LOINC terms
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: no
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

# Introduction

The purpose of this preliminary analysis is to investigate likely molecular
biomarkers.

* [LOINC](https://loinc.org/) _(Logical Observation Identifiers Names and Codes)_
is developed and maintained by the [Regenstrief Institute](https://www.regenstrief.org/)  and widely used for encoding clinical lab records in EHR systems.
* [Oracle HealthFacts](https://www.oracle.com/health/population-health/real-world-data/)
contains over 42 million US patient records from the previous two decades,
with lab results encoded with LOINC. Formerly from Cerner, acquired by Oracle in 2021.

## Biomarkers

The [FDA Biomarker Qualification Program](https://www.fda.gov/drugs/biomarker-qualification-program/about-biomarkers-and-qualification), via its Biomarkers, EndpointS and other Tools [BEST](https://www.ncbi.nlm.nih.gov/books/NBK326791/) glossary, defines biomarker thus: "[A] defined characteristic that is measured as an indicator of normal biological processes, pathogenic processes, or responses to an exposure or intervention, including therapeutic interventions. Molecular, histologic, radiographic, or physiologic characteristics are types of biomarkers." This analysis is concerned with ___molecular___ biomarkers only.

# Workflow

* Download [LOINC](https://loinc.org/) codes.
* Build local LOINC PostgreSql db. Query for chemical biomarkers, to include proteins.
* Query Oracle HealthFacts 2019 (licensed at UNM) to find encounter_id and patient_id counts for all LOINC codes, for clinically relevant biomarkers (not just theoretical).
* Generate initial list of clinically relevant molecular biomarker candidates.

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r setup, echo=FALSE, message=FALSE}
library(readr)
library(stringr)
library(data.table)
library(plotly, quietly=T)
```

# LOINC version

```{r}
LOINC_RELEASE <- trimws(read_file("LATEST_RELEASE_LOINC.txt"))
DATADIR <- sprintf("loinc_data/v%s", LOINC_RELEASE)
sprintf("LOINC_RELEASE: %s", LOINC_RELEASE)
```

# LOINC codes for chemical tests

```{r}
loinc_file <- sprintf("%s/loinc_chem_names.tsv", DATADIR)
sprintf("LOINC codes for chemical tests: %s", loinc_file)
loinc_chem <- read_delim(loinc_file, "\t", escape_double=F)
setDT(loinc_chem)
sprintf("LOINC codes: %d", loinc_chem[, uniqueN(loinc_num)])
```

# Oracle HealthFacts 2019 counts of encounter\_ids and patient\_ids for all lab procedures, with LOINC codes.

```{r}
hf_loinc_counts_file <- "oracle_hf_data/hf_lab_loinc_counts_OUT.tsv"
sprintf("Oracle HealthFacts 2019 counts for labs, with LOINC codes: %s", hf_loinc_counts_file)
hf_loinc_counts <- read_delim(hf_loinc_counts_file, "\t")
setDT(hf_loinc_counts)
sprintf("Lab procedure IDs: %d; LOINC codes: %d", hf_loinc_counts[, uniqueN(lab_procedure_id)], hf_loinc_counts[, uniqueN(loinc_code)])
```

## Group lab procedures into list; aggregate on LOINC codes.

Some different labs have same LOINC codes. Discarding lab procedures IDs.

```{r}
hf_loinc_counts <- hf_loinc_counts[, .(
	lab_procedure_names = paste(c(lab_procedure_name), collapse="; "),
	lab_procedure_mnemonics = paste(c(lab_procedure_mnemonic), collapse="; "),
	lab_super_groups = paste(base::unique(c(lab_super_group)), collapse="; "),
	encounter_id_count = sum(encounter_id_count),
	patient_id_count = sum(patient_id_count)
	), by=c("loinc_code")]

```

# LOINC proteins in HF encounters

```{r}
protein_hr_loincs_component <- merge(protein_NER_component, hf_loinc_counts, by.x="DocName", by.y="loinc_code", all=F)
protein_hr_loincs_relatedname <- merge(protein_NER_relatedname, hf_loinc_counts, by.x="DocName", by.y="loinc_code", all=F)
protein_hr_loincs <- rbind(protein_hr_loincs_component, protein_hr_loincs_relatedname)
protein_hr_loincs <- unique(protein_hr_loincs)
sprintf("LOINC codes for proteins: %d", protein_hr_loincs[, uniqueN(DocName)])
sprintf("Protein names: %d", protein_hr_loincs[, uniqueN(EntityText)])
sprintf("GeneOrProtein IDs: %d", protein_hr_loincs[, uniqueN(ResolvedForm)])
```


# Top occurring LOINCs:

```{r}
protein_hr_loincs_out <- protein_hr_loincs_out[order(-encounter_id_count)]
write_delim(protein_hr_loincs_out, "oracle_hf_data/biomarkers_loinc_hf_out.tsv", delim="\t")
knitr::kable(protein_hr_loincs_out[1:min(50, nrow(protein_hr_loincs_out))], caption=sprintf("LOINCs ranked by occurence: %d LOINCs, %d geneOrProteinIds", protein_hr_loincs_out[, uniqueN(loinc_code)], protein_hr_loincs_out[, uniqueN(geneOrProteinId)]))
```
