---
title: 'Cerner HealthFacts: Laboratory molecular biomarker results analysis'
output:
  html_document:
    df_print: paged
---

```{r include=F}
library(readr)
library(data.table)
library(plotly)
```

## Read data file for sample, representative year, due to size constraints.

```{r}
y <- 2016
fpath_in <- sprintf("cerner_hf_data/hf_labs-selected_results_OUT_%d.tsv.gz", y)
#hf_results <- read_delim(fpath_in, delim = "\t", col_types = cols(.default = col_character(), admitted_dt_tm = col_datetime(), numeric_result = col_double(), normal_range_low = col_double(), normal_range_high = col_double()))
hf_results <- read_delim("cerner_hf_data/hf_labs-selected_results_OUT_2016.tsv.gz", delim = "\t", escape_double = FALSE,  trim_ws = TRUE, col_types = cols(encounter_id = col_character()))
setDT(hf_results)
hf_results <- hf_results[, numeric_result := as.numeric(numeric_result)]
hf_results <- hf_results[, normal_range_low := as.numeric(normal_range_low)]
hf_results <- hf_results[, normal_range_high := as.numeric(normal_range_high)]
hf_results <- hf_results[, admitted_dt_tm := as.Date(admitted_dt_tm)]
head(hf_results)
```

```{r}
message(sprintf("Columns: %s", paste(names(hf_results), collapse=", ")) )
```

```{r}
message(sprintf("ENCOUNTERs: %s", hf_results[, uniqueN(encounter_id)]))
message(sprintf("ACCESSIONs: %s", hf_results[, uniqueN(accession)]))
message(sprintf("LAB_PROCEDURE_IDs: %s", hf_results[, uniqueN(lab_procedure_id)]))
message(sprintf("LOINC_CODEs: %s", hf_results[, uniqueN(loinc_code)]))
message(sprintf("RESULT_UNIT_IDs: %s", hf_results[, uniqueN(result_units_id)]))
```
Map LOINC codes to GeneOrProtein IDs.

```{r}
loinc2gp <- read_delim("loinc_data/loinc2gp.tsv", delim="\t", escape_double=F,  trim_ws=T, col_types=cols(loinc_code = col_character()))
setDT(loinc2gp)
hf_results <- merge(hf_results, loinc2gp, by="loinc_code", all.x=T, all.y=F)
message(sprintf("GeneOrProtein_IDs: %s", hf_results[, uniqueN(geneOrProteinId)]))
```


```{r}
loinc2gp_results <- unique(hf_results[!is.na(geneOrProteinId), .(lab_procedure_id, lab_procedure_mnemonic, lab_procedure_name, geneOrProteinId)])
knitr::kable(loinc2gp_results[, .(lab_procedure_id, lab_procedure_mnemonic, geneOrProteinId)][order(lab_procedure_mnemonic)])
```

## Find and focus on results for PSA (prostate specific antigen)

```{r}
psa_ids <- unique(hf_results[grepl("PSA", lab_procedure_name, ignore.case=T), lab_procedure_id])
knitr::kable(unique(hf_results[lab_procedure_id %in% psa_ids, .(lab_procedure_id, lab_procedure_mnemonic, lab_procedure_name)]))
```

```{r}
hf_results[unit_display == "NULL", unit_display := NA]
hf_results[unit_desc == "NULL", unit_desc := NA]
knitr::kable(hf_results[lab_procedure_id == "1237" & !is.na(numeric_result) & !is.na(unit_display), .(lab_procedure_id, lab_procedure_mnemonic, numeric_result, unit_display, unit_desc)][sample(1:1e4, 50)])
```

## PSA units analysis

```{r}
psa_units <- unique(hf_results[lab_procedure_id == "1237" & !is.na(numeric_result), .(unit_display, unit_desc)])
tbl <- table(hf_results[lab_procedure_id == "1237" & !is.na(numeric_result), unit_display])
psa_units <- merge(psa_units, data.table(tbl), by.x="unit_display", by.y="V1")
psa_units <- psa_units[order(-N)]
knitr::kable(psa_units)
psa_units_preferred <- psa_units[["unit_display"]][1]
message(sprintf("PSA units (preferred): %s", psa_units_preferred))
```


## PSA value analysis

Keep only results with predominant units. To do: harmonize via unit conversion as possible.

```{r}
hf_results_psa <- hf_results[lab_procedure_id == "1237" & !is.na(numeric_result) & unit_display == psa_units_preferred]
sprintf("Lab procedure: %s, \"%s\"; values: %d; units: %s; min: %.1f; max: %.1f", hf_results_psa[, lab_procedure_mnemonic][1], hf_results_psa[, lab_procedure_name][1], hf_results_psa[, .N], unique(hf_results_psa[, unit_display]), min(hf_results_psa[, numeric_result]), max(hf_results_psa[, numeric_result]))
```

## Consider quantiles, and remove outlier values as likely spurious.

```{r}
q <- quantile(hf_results_psa[, numeric_result], probs = c(seq(0, .9, .1), seq(.91, .99, .01), seq(.991, 1, .001)))
psa_sane_max <- q["99%"]
hf_results_psa <- hf_results_psa[numeric_result <= psa_sane_max & numeric_result > 0]
data.table(percentiles = names(q), values = q)
```

```{r}
fig <- plot_ly(alpha = 0.8) %>% 
	add_histogram(x = hf_results_psa[, numeric_result], name = "PSA", marker = list(color = "deeppink"))
fig
```

## Merge patient metadata for stratification.

Ignore non-gendered in this study.

```{r}
ptdata <- read_delim(sprintf("cerner_hf_data/hf_labs-selected_pt_OUT_ptdata_%d.tsv.gz", y), delim = "\t", col_types = cols(.default = col_character(), admitted_year = col_integer(), age_in_years = col_integer()))
setDT(ptdata)
ptdata <- ptdata[gender == "Male" | gender == "Female"]
#tbl <- table(ptdata[, .(race, gender)])
#tbl <- data.table(as.data.frame.matrix(tbl), keep.rownames = T)
#names(tbl) <- c("Race", "Female", "Male")
#tbl <- rbind(tbl, list("TOTAL", tbl[, sum(Female)], tbl[, sum(Male)]))
#tbl[, Fpct := 100 * Female / tbl[Race != "TOTAL", sum(Female)]]
#tbl[, Mpct := 100 * Male / tbl[Race != "TOTAL", sum(Male)]]
#tbl <- tbl[, .(Race, Female, Fpct, Male, Mpct)]
#write_delim(tbl, sprintf("cerner_hf_data/hf_labs-selected_pt_OUT_ptdata_%d_race-gender-counts-table.tsv", y), delim = "\t")
#knitr::kable(tbl, digits = 1, format = "html", format.args = list(big.mark = ","))
```

Create "Agerange" column (decades).

```{r}
ptdata[, agerange := fifelse(age_in_years<10, "0-9", 
                       fifelse(age_in_years<20, "10-19", 
                       fifelse(age_in_years<30, "20-29",
                       fifelse(age_in_years<40, "30-39",
                       fifelse(age_in_years<50, "40-49",
                       fifelse(age_in_years<60, "50-59",
                       fifelse(age_in_years<70, "60-69",
                       fifelse(age_in_years<80, "70-79",
                       fifelse(age_in_years<90, "80-89",
                       fifelse(age_in_years<100, "90-99", "100+"
                               ))))))))))]
```


```{r}
tbl <- table(ptdata[, .(agerange, gender)])
tbl <- data.table(as.data.frame.matrix(tbl), keep.rownames = T)
names(tbl) <- c("Agerange", "Female", "Male")
tbl <- rbind(tbl, list("TOTAL", tbl[, sum(Female)], tbl[, sum(Male)]))
tbl[, Fpct := 100 * Female / tbl[Agerange != "TOTAL", sum(Female)]]
tbl[, Mpct := 100 * Male / tbl[Agerange != "TOTAL", sum(Male)]]
tbl <- tbl[, .(Agerange, Female, Fpct, Male, Mpct)]
write_delim(tbl, sprintf("cerner_hf_data/hf_labs-selected_pt_OUT_ptdata_%d_agerange-gender-counts-table.tsv", y), delim = "\t")
knitr::kable(tbl, digits = 1, format = "html", format.args = list(big.mark = ","))
```


```{r}
eid2pid <- read_delim(sprintf("cerner_hf_data/hf_labs-selected_pt_OUT_eid2pid_%d.tsv.gz", y), delim = "\t", col_types = cols(.default = col_character()))
setDT(eid2pid)
```

## Consider only male, stratify by age

```{r}
hf_results_psa <- merge(hf_results_psa, eid2pid, by="encounter_id", all.x=T, all.y=F)
hf_results_psa <- merge(hf_results_psa, ptdata[, .(patient_id, gender, agerange)], by="patient_id", all.x=T, all.y=F)
hf_results_psa <- hf_results_psa[gender == "Male" & !is.na(patient_id) & numeric_result <= 8.0]
```

## Histograms by age group

```{r}
fig <- plot_ly(alpha = 0.8) %>% 
	add_histogram(x = hf_results_psa[agerange %in% c("0-9", "10-19", "21-29", "31-39"), numeric_result], name = "Under40yrs") %>%
	add_histogram(x = hf_results_psa[agerange == "40-49", numeric_result], name = "40-49yrs") %>%
	add_histogram(x = hf_results_psa[agerange == "50-59", numeric_result], name = "50-59yrs") %>%
	add_histogram(x = hf_results_psa[agerange == "60-69", numeric_result], name = "60-69yrs") %>%
	add_histogram(x = hf_results_psa[agerange == "70-79", numeric_result], name = "70-79yrs") %>%
	add_histogram(x = hf_results_psa[agerange %in% c("80-89", "90-99", "100+"), numeric_result], name = "80+yrs") %>%
layout(barmode = "stack",
				 title = sprintf("Histogram of PSA by age-range, male only (year: %d)", y),
         xaxis = list(title = "PSA (ng/mL)", zeroline = F),
         yaxis = list(title = "Encounters", zeroline = F),
			 legend = list(x = 0.8, y = 0.9))
fig
```

Compute sample mean and variance, for each lab, and each strata.

```{r}
message(sprintf("PSA units: %s; mean: %.2f; variance: %.2f",
								psa_units_preferred,
								mean(hf_results_psa[["numeric_result"]], na.rm=T),
								var(hf_results_psa[["numeric_result"]], na.rm=T)
								))
```

Add patient sex and agerange to results dataset.
Remove patient_id, encounter_id, age, and race to enhance deidentification.
Remove other unused columns to save space.
Remove ptdata rows with duplicate patient_id (though different data)

```{r}
hf_results <- merge(hf_results, eid2pid, by="encounter_id", all.x=T, all.y=F)
ptdata <- ptdata[!duplicated(ptdata[["patient_id"]])]
hf_results <- hf_results[!is.na(patient_id)]
hf_results <- merge(hf_results, ptdata[, .(patient_id, gender, agerange)], by="patient_id", all.x=T, all.y=F)
hf_results[, ':='(encounter_id=NULL, patient_id=NULL, lab_procedure_id=NULL, accession=NULL, admitted_dt_tm=NULL, normal_range_low=NULL, normal_range_high=NULL)]
message(sprintf("OUTPUT: lines: %d; LOINC_CODEs: %d; GENEORPROTEIN_IDs: %d", nrow(hf_results), hf_results[, uniqueN(loinc_code)], hf_results[, uniqueN(geneOrProteinId)]))
ofile <- "cerner_hf_data/hf_labs-selected_results_sex-agerange_OUT.tsv.gz"
message(sprintf("OUTPUT: writing to %s", nrow(hf_results), ofile))
write_delim(hf_results, ofile, delim="\t")
```

