---
title: Human protein biomarkers, with clinical relevance inferred from Cerner RealWorldData
  via LOINC terms
author: "Jeremy Yang"
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r setup, echo=FALSE, message=FALSE}
library(readr)
library(data.table)
library(plotly, quietly=T)
```

# LOINC codes for chemical tests

```{r}
loinc_file <- "loinc_data/loinc_chem_names.tsv"
sprintf("LOINC codes for chemical tests: %s", loinc_file)
loinc_chem <- read_delim(loinc_file, "\t", escape_double=F)
setDT(loinc_chem)
sprintf("LOINC codes: %d", loinc_chem[, uniqueN(loinc_num)])
```

# NextMove LeadMine NER output using GeneAndProtein dictionary.

NER = named entity recognition, a form of text-mining.

## NER on LOINC field: "component".

```{r}
protein_NER_component_file <- "loinc_data/loinc_chem_names_2_NM_CFDictGeneAndProtein_leadmine.tsv"
sprintf("NextMove NER for LOINC-component: %s", protein_NER_component_file)
protein_NER_component <- read_delim(protein_NER_component_file, "\t")
setDT(protein_NER_component)
sprintf("LOINC codes mapped via component text to NER proteins: %d", protein_NER_component[, uniqueN(DocName)])
```

## NER on LOINC field: "relatednames2".

```{r}
protein_NER_relatednames_file <-  "loinc_data/loinc_chem_names_6_NM_CFDictGeneAndProtein_leadmine.tsv"
sprintf("NextMove NER for LOINC-relatednames: %s", protein_NER_relatednames_file)
protein_NER_relatednames <-  read_delim(protein_NER_relatednames_file, "\t")
setDT(protein_NER_relatednames)
sprintf("LOINC codes mapped via relatednames2 text to NER proteins: %d", protein_NER_relatednames[, uniqueN(DocName)])
```

# Cerner RWD counts

Cerner Real World Data 2023 counts of encounter\_ids and patient\_ids for all LOINC codes.

```{r}
rwd_loinc_counts <- read_csv("cerner_rwd_data/base_counts.csv")
setDT(rwd_loinc_counts)
sprintf("LOINC codes in RWD encounters: %d", rwd_loinc_counts[, uniqueN(loinc_code)])
```


## Group lab procedures into list; aggregate on LOINC codes.

Some different labs have same LOINC codes.

```{r}
rwd_loinc_counts <- rwd_loinc_counts[, .(
	mnemonics = paste(base::unique(c(mnemonic)), collapse="; "),
	loinc_classes = paste(base::unique(c(loincclass)), collapse="; "),
	years = paste(sort(base::unique(c(yr))), collapse="; "),
	encounter_id_count = sum(count_encs),
	patient_id_count = sum(count_pts)
	), by=c("loinc_code")]

```

# LOINC proteins in RWD encounters

```{r}
protein_hr_loincs_component <- merge(protein_NER_component, rwd_loinc_counts, by.x="DocName", by.y="loinc_code", all=F)
protein_hr_loincs_relatednames <- merge(protein_NER_relatednames, rwd_loinc_counts, by.x="DocName", by.y="loinc_code", all=F)
protein_hr_loincs <- rbind(protein_hr_loincs_component, protein_hr_loincs_relatednames)
sprintf("LOINC codes for proteins in RWD encounters: %d", protein_hr_loincs[, uniqueN(DocName)])
sprintf("LOINC protein names in RWD encounters: %d", protein_hr_loincs[, uniqueN(EntityText)])
```

## Group protein synonyms into list; aggregate on LOINC codes.

```{r}
protein_hr_loincs_out <- unique(protein_hr_loincs[, .(loinc_code=DocName, mnemonics, protein_name=EntityText, encounter_id_count, patient_id_count)])
protein_hr_loincs_out <- protein_hr_loincs_out[, .(mnemonics,
protein_names = paste(c(protein_name), collapse="; ")), by=c("loinc_code", "encounter_id_count", "patient_id_count")]
protein_hr_loincs_out <- unique(protein_hr_loincs_out)
setcolorder(protein_hr_loincs_out, c("loinc_code", "mnemonics", "protein_names", "encounter_id_count", "patient_id_count"))
```

# Top occurring LOINCs:

```{r}
protein_hr_loincs_out <- protein_hr_loincs_out[order(-encounter_id_count)]
write_delim(protein_hr_loincs_out, "cerner_hf_data/biomarkers_loinc_hf_out.tsv", delim="\t")
knitr::kable(protein_hr_loincs_out[1:50], caption="Top occurring LOINCs")
```

