---
title: Biomarkers from FDA CLIA approved and classified lab tests
author: "Jeremy Yang"
output:
  html_document: 
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

FDA CLIA files are downloaded [here](https://www.fda.gov/medical-devices/medical-device-databases/clinical-laboratory-improvement-amendments-download-data).

Regarding the document numbers, "either the 510(k), PMA, HDE or 510(k) exempt number", it appears the codes starting with 'K' are 510(k)s, the codes starting with 'P' are PMAs (Pre-market approval), the codes starting with 'CR' are document numbers.

The analytes are an explicit, structured field, with an ID and name. But what are the tests for? Where are the clinical conditions which the tests are supposed to indicate? Do we need to process the referenced documents?

Examples:

 * [CR190560](https://www.accessdata.fda.gov/Scripts/cdrh/cfdocs/cfClia/Results.cfm?start_search=1&Document_Number=CR190560&SortColumn=ded)
 * [K124056](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfPMN/pmn.cfm?ID=K124056)
 * [P000021S045](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfpma/pma.cfm?id=P000021S045)

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r setup, echo=FALSE, message=FALSE}
library(readr)
library(stringr)
library(data.table)
library(plotly, quietly=T)
```

# FDA CLIA Detail file

```{r}
clia_detail_file <- "fda_clia_data/clia_detail.tsv"
sprintf("FDA CLIA DETAIL file: %s", clia_detail_file)
clia_detail <- read_delim(clia_detail_file, "\t", escape_double=F)
setDT(clia_detail)
sprintf("DOCUMENT_NUMBERs: %d; TEST_SYSTEM_IDs: %d; ANALYTE_IDs: %d", clia_detail[, uniqueN(DOCUMENT_NUMBER)], clia_detail[, uniqueN(TEST_SYSTEM_ID)], clia_detail[, uniqueN(ANALYTE_ID)])
```

# NextMove LeadMine NER output using GeneAndProtein dictionary.

```{r}
clia_detail_ner_gp_file <- "fda_clia_data/clia_detail_gene_or_protein_leadmine.tsv"
sprintf("CLIA DETAIL Gene-Protein-NER file: %s", clia_detail_ner_gp_file)
clia_detail_ner_gp <- read_delim(clia_detail_ner_gp_file, "\t", escape_double=F)
setDT(clia_detail_ner_gp)
clia_detail_ner_gp <- clia_detail_ner_gp[, .(AnalyteId = DocName, OriginalText, EntityText, GeneOrProteinId = ResolvedForm)]
clia_detail_ner_gp <- clia_detail_ner_gp[, GeneOrProteinId := str_replace(GeneOrProteinId, "^([A-Z]\\d+)$", "UNIPROT:\\1")]
sprintf("Analytes resolved to standard gene-or-protein IDs: %d / %d", clia_detail_ner_gp[!is.na(GeneOrProteinId), uniqueN(AnalyteId)], clia_detail_ner_gp[, uniqueN(AnalyteId)])
sprintf("Analytes resolved to HGNC IDs: %d / %d", clia_detail_ner_gp[!is.na(GeneOrProteinId) & grepl("^HGNC", GeneOrProteinId), uniqueN(AnalyteId)], clia_detail_ner_gp[, uniqueN(AnalyteId)])
sprintf("Analytes resolved to UniProt IDs: %d / %d", clia_detail_ner_gp[!is.na(GeneOrProteinId) & grepl("^UNIPROT:", GeneOrProteinId), uniqueN(AnalyteId)], clia_detail_ner_gp[, uniqueN(AnalyteId)])
sprintf("Analytes resolved to Entrez (gene?) IDs: %d / %d", clia_detail_ner_gp[!is.na(GeneOrProteinId) & grepl("^\\d+$", GeneOrProteinId), uniqueN(AnalyteId)], clia_detail_ner_gp[, uniqueN(AnalyteId)])
clia_detail_ner_gp <- clia_detail_ner_gp[!is.na(GeneOrProteinId)] #Delete unmapped records
clia_detail_ner_gp <- clia_detail_ner_gp[!grepl("^\\d+$", GeneOrProteinId)] #Ignore (too few) NCBI IDs
clia_detail_ner_gp <- unique(clia_detail_ner_gp[order(AnalyteId)])
knitr::kable(clia_detail_ner_gp[sample(.N, 20)], caption="Sample of CLIA NER Genes")
```

## Documents, Tests, and Analytes

Note that DocNum to Test to Analyte relationships are many-to-many.

```{r}
clia_doc2test2analyte <- unique(clia_detail[, .(DOCUMENT_NUMBER, TEST_SYSTEM_ID, ANALYTE_ID, ANALYTE_NAME, DATE_EFFECTIVE)])[order(DOCUMENT_NUMBER, ANALYTE_ID, TEST_SYSTEM_ID, DATE_EFFECTIVE)]
knitr::kable(clia_doc2test2analyte[grepl("prostat.*antigen", ANALYTE_NAME, ignore.case=T)][sample(.N, 20)], caption = "Sample of 'prostat.*antigen' matches")
```

## Merge NER with analyte metadata (names), save file

Analyte IDs can be mapped to CLIA documents via detail file.

```{r}
clia_detail_ner_gp <- merge(clia_detail_ner_gp, unique(clia_detail[, .(ANALYTE_ID, ANALYTE_NAME)]), by.x="AnalyteId", by.y="ANALYTE_ID")
clia_detail_ner_gp <- clia_detail_ner_gp[, .(AnalyteId, AnalyteName = ANALYTE_NAME, OriginalText, EntityText, GeneOrProteinId)]
clia_detail_ner_gp <- clia_detail_ner_gp[, .(OriginalText = first(OriginalText), EntityText = first(EntityText)), by = .(AnalyteId, AnalyteName, GeneOrProteinId)]
clia_detail_ner_gp_outfile <- "output/clia_detail_GeneOrProtein.tsv"
write_delim(clia_detail_ner_gp, clia_detail_ner_gp_outfile, delim="\t")
```

# Generate mapping file of GeneOrProtein analytes to CLIA documents and tests.

```{r}
clia_doc2test2analyte_this <- unique(merge(clia_doc2test2analyte, clia_detail_ner_gp[, .(AnalyteId)], by.x="ANALYTE_ID", by.y="AnalyteId", all.x=F, all.y=F, sort=T))
clia_doc2test2analyte_outfile <- "output/clia_doc2test2analyte_gp.tsv"
write_delim(clia_doc2test2analyte_this, clia_doc2test2analyte_outfile, delim="\t")
```


