---
title: 'Oracle HealthFacts: Laboratory molecular biomarker results analysis: Troponin biomarkers'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r echo=F, include=F}
library(readr)
library(data.table)
library(plotly)
library(kableExtra)
```

## Read data file.


```{r echo=F}
ifile <- "oracle_hf_data/hf_f_lab_troponin_2018_out.tsv"
hf_lab <- read_delim(ifile, "\t", escape_double=F, show_col_types = F)
setDT(hf_lab)
sprintf("Columns: %s", paste(names(hf_lab), collapse=", "))
sprintf("Patients: %s; Encounters: %s", hf_lab[, uniqueN(patient_id)], hf_lab[, uniqueN(encounter_id)])
```

## Tests in this dataset.
```{r echo=F}
hf_lab[, lab_procedure_mnemonic := as.factor(lab_procedure_mnemonic)]
tbl <- data.table(table(hf_lab[, lab_procedure_mnemonic]))
names(tbl) <- c("Lab", "N")
tbl <- tbl[order(-N)]
data.table(tbl) %>% 
  kbl(caption = "Lab tests in dataset, with value counts", booktabs=T) %>%
  kable_styling(c("striped", "hover"), full_width = F, position = "left")
```

## Define biomarker[s] of interest.

```{r echo=F}
boi <- list("Troponin I" = NULL, "Troponin T S/Pl" = NULL)
for (lpm in names(boi)) {
  message(sprintf("Biomarker of interest: \"%s\"", lpm))
}
```

## Describe dataset[s] for biomarker[s] of interest.

```{r echo=F}
report_dt <- data.table(i = 1:length(boi), lab = "", valcount = NaN, all_units = "", pref_units = "", pref_units_valcount = NaN, mean = NaN, median = NaN, sd = NaN)
i <- 0L
for (lpm in names(boi)) {
  hf_lab_this <- hf_lab[lab_procedure_mnemonic == lpm]
  i <- i + 1
  report_dt$lab[i] <- paste(unique(hf_lab_this[["lab_procedure_mnemonic"]]), collapse=", ")
  report_dt$valcount[i] <- hf_lab_this[, .N]
  tbl <- data.table(table(hf_lab_this[["unit_display"]]))[order(-N)]
  units_preferred <- tbl[["V1"]][1]
  report_dt$all_units[i] <- paste(unique(hf_lab_this[["unit_display"]]), collapse=", ")
  report_dt$pref_units[i] <- units_preferred
  report_dt$pref_units_valcount[i] <- tbl[["N"]][1]
  hf_lab_this <- hf_lab_this[unit_display == units_preferred]
  report_dt$mean[i] <- mean(hf_lab_this[["numeric_result"]], na.rm=T)
  report_dt$median[i] <- median(hf_lab_this[["numeric_result"]], na.rm=T)
  report_dt$sd[i] <- sd(hf_lab_this[["numeric_result"]], na.rm=T)
  boi[[lpm]] <- hf_lab_this
}
report_dt[, .(lab, pref_units, pref_units_valcount, mean, median, sd)] %>% 
  kbl(caption = sprintf("Summary of data for biomarker[s] of interest (%d)", length(boi)), booktabs=T, digits=2, align = "lccccc", col.names = c("lab", "units", "N", "mean", "median", "sd")) %>%
    kable_styling(c("striped", "hover"), full_width = F, position = "left")
```
```{r echo=F, results='asis'}
for (lpm in names(boi)) {
  #hf_lab_this <- hf_lab[lab_procedure_mnemonic == lpm]
  hf_lab_this <- boi[[lpm]]
  values <- hf_lab_this[["numeric_result"]]
  message(sprintf("%s; N = %d; min = %.4f; max = %.4f; mean = %.4f; median = %.4f; sd = %.4f", lpm, length(values), min(values), max(values), mean(values), median(values), sd(values)))
  qtl <- quantile(values, probs = seq(0, 1, .1))
  print(
  data.table(percentiles = names(qtl), values = qtl) %>%
  kbl(caption = sprintf("%s: distribution percentiles", lpm), booktabs=T, align = "cc") %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
  )
}
```

# Remove extreme and likely non-healthy or spurious values.

Although exclusion criteria are designed to filter for healthy patients with normal physiological lab values, removing extreme values can further filter to the same purpose.

```{r echo=F}
for (lpm in names(boi)) {
  hf_lab_this <- boi[[lpm]]
  values <- hf_lab_this[, numeric_result]
  n_before <- length(values)
  qtl <- quantile(values, probs = seq(0, 1, .05))
  hf_lab_this <- hf_lab_this[numeric_result>qtl[["10%"]] & numeric_result<qtl[["90%"]]]
  values <- hf_lab_this[, numeric_result]
  message(sprintf("%s: Extreme values removed: %d / %d (%.2f%%)", lpm, n_before-length(values), n_before, 100 * (n_before-length(values))/n_before))
  boi[[lpm]] <- hf_lab_this
  message(sprintf("%s; N = %d; min = %.4f; max = %.4f; mean = %.4f; median = %.4f; sd = %.4f", lpm, length(values), min(values), max(values), mean(values), median(values), sd(values)))
  qtl <- quantile(values, probs = seq(0, 1, .1))
  data.table(percentiles = names(qtl), values = qtl) %>% 
    kbl(caption = sprintf("%s: distribution percentiles", lpm), booktabs=T) %>%
    kable_styling(c("striped", "hover"), full_width = F, position="left")
}
```


```{r echo=F}
plots <- list()
i <- 0L
for (lpm in names(boi)) {
  values <- boi[[lpm]][, numeric_result]
  boi[[lpm]][, unit_display := as.factor(unit_display)]
  units_preferred <- levels(boi[[lpm]][, unit_display])[1]
  plotname <- sprintf("%s (%s)<br>N = %d", lpm, units_preferred, length(values))
  p <- plot_ly(type="histogram", x = values, name=plotname, xbins = list(size = 0.01)) %>%
    layout(annotations=list(text = plotname, x=0.5, y=0.7, xref="paper", yref="paper",
                yanchor="bottom", xanchor="center", align="center",
                font=list(family="Courier New", size=12, color="#222222"), showarrow=F))
  i <- i + 1
  plots[[i]] <- p
}
subplot(plots, nrows=1, shareX=F, shareY=T, titleX=T, titleY=T, margin=0.03) %>%
  layout(title=sprintf("Distributions: %s", paste(collapse=", ", names(boi))),
         margin = list(t = 60),
         font = list(family = "Arial", size = 14), showlegend = T)
```


```{r echo=F}
ifile <- "oracle_hf_data/hf_f_lab_troponin_2018_patients_out.tsv"
hf_pt <- read_delim(ifile, "\t", escape_double=F, show_col_types=F)
setDT(hf_pt)
sprintf("Columns: %s", paste(names(hf_pt), collapse=", "))
```

## Patient cohort descriptive statistics

### Remove patients with unknown gender or age

```{r echo=F}
sprintf("Excluding patients with unknown gender: %d / %d (%.3f%%) excluded",  
        hf_pt[gender != "Female" & gender != "Male", .N],
        hf_pt[, .N], 100 * hf_pt[gender != "Female" & gender != "Male", .N] / hf_pt[, .N])
sprintf("Excluding patients with unknown agegroup: %d / %d (%.3f%%) excluded", 
        hf_pt[agegroup == "Unknown", .N], 
        hf_pt[, .N], 100 * hf_pt[agegroup == "Unknown", .N] / hf_pt[, .N])
hf_pt <- hf_pt[gender == "Female" | gender == "Male"]
hf_pt <- hf_pt[agegroup != "Unknown"]
```

### Before and after applying exclusion criteria (diagnosis related to heart disease).

```{r echo=F}
sprintf("Patients: %d; Not-excluded: %d (%.1f%%)", 
        hf_pt[, .N], 
        hf_pt[exclusion_flag==F, .N], 
        100 * hf_pt[exclusion_flag==F, .N] / hf_pt[, .N])
tbl <- data.table(table(hf_pt[, gender]))
names(tbl) <- c("Gender", "n_pt")
tbl <- tbl[order(-n_pt)]
tbl %>%
  kbl(caption = sprintf("Labs: (%s) cohort, pre-exclusion, Gender distribution", paste(collapse=", ", names(boi))), booktabs=T) %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
```

```{r echo=F}
hf_pt <- hf_pt[exclusion_flag==F]
tbl <- data.table(table(hf_pt[, gender]))
names(tbl) <- c("Gender", "n_pt")
tbl <- tbl[order(-n_pt)]
tbl %>%
  kbl(caption = sprintf("Labs: (%s) cohort, Gender distribution", paste(names(boi), collapse=", ")), booktabs=T) %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
```
## Patient type

Note that patient type depends on the encounter, so a single patient can be multiple patient types for different encounters.

```{r echo=F}
tbl <- table(hf_pt[, patient_type_desc])
tbl <- data.table(tbl)
names(tbl) <- c("Patient_type", "n_pt")
tbl <- tbl[order(-n_pt)]
tbl %>%
  kbl(caption = sprintf("Labs: (%s) cohort, Patient_type distribution", paste(names(boi), collapse=", ")), booktabs=T) %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
```

## Race

```{r echo=F}
tbl <- table(hf_pt[, race])
tbl <- data.table(tbl)
names(tbl) <- c("Race", "n_pt")
tbl <- tbl[order(-n_pt)]
tbl %>%
  kbl(caption = sprintf("Labs: (%s) cohort, Race distribution", paste(names(boi), collapse=", ")), booktabs=T) %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
```

## Age distribution

```{r echo=F}
values <- hf_pt[!is.na(age_mid_year), age_mid_year]
sprintf("Age_mid_year: N = %d; min = %.1f; max = %.1f; mean = %.1f; median = %.1f; sd = %.1f", length(values), min(values), max(values), mean(values), median(values), sd(values))
qtl <- quantile(values, probs = seq(0, 1, .1))
data.table(percentiles = names(qtl), values = qtl) %>%
  kbl(caption = sprintf("Labs: (%s) cohort, Age_mid_year: distribution percentiles", paste(names(boi), collapse=", ")), booktabs=T, digits=1) %>%
  kable_styling(c("striped", "hover"), full_width = F, position="left")
```



```{r echo=F}
fig <- plot_ly(alpha = 0.8, xbins = list(size=5)) %>% 
	add_histogram(x = hf_pt[gender == "Female", age_mid_year], name = "Female", marker = list(color = "deeppink")) %>%
	add_histogram(x = hf_pt[gender == "Male", age_mid_year], name = "Male", marker = list(color = "skyblue")) %>%
	layout(barmode = "group",
				 title = sprintf("Labs: (%s) cohort: Histogram of patient count by age, gender", paste(names(boi), collapse=", ")),
         xaxis = list(title = "Age", zeroline = F),
         yaxis = list(title = "Patients", zeroline = F),
				 legend = list(x = .2, y = .7))
fig
```

## Stratified statistics and  distributions

Initially, for sex and age-group.
First merge lab value table[s] with stratification variables.

```{r echo=F}
for (lpm in names(boi)) {
  boi[[lpm]] <- merge(boi[[lpm]], hf_pt[, .(patient_id, gender, race, age_mid_year, agegroup, exclusion_flag)], by = "patient_id", all.x=F, all.y=F)
}
#hf_lab_pt_this <- merge(hf_lab_this, hf_pt[, .(patient_id, gender, race, age_mid_year, agegroup, exclusion_flag)], by = "patient_id", all.x=F, all.y=F)
```


```{r echo=F}
for (lpm in names(boi)) {
  hf_lab_pt_this <- boi[[lpm]]
  hf_lab_pt_this[, gender := as.factor(gender)]
  hf_lab_pt_this[, agegroup := as.factor(agegroup)]
  hf_lab_pt_this[, unit_display := as.factor(unit_display)]
  n_agegroup <- length(levels(hf_lab_pt_this[, agegroup]))
  report_stratstat <- data.table(i = 1:(2*n_agegroup), sex = "", agegroup = "", valcount = NaN, units = "", mean = NaN, median = NaN, sd = NaN)
  i <- 0L
  for (a in levels(hf_lab_pt_this[, agegroup])) {
    for (g in levels(hf_lab_pt_this[, gender])) {
      i <- i + 1
      ptid_this <- hf_lab_pt_this[agegroup == a & gender == g, patient_id]
      #message(sprintf("Agegroup: %s; Gender: %s; N = %d", a, g, length(ptid_this)))
      values <- hf_lab_pt_this[agegroup == a & gender == g, numeric_result]
      units <- hf_lab_pt_this[agegroup ==a & gender == g, unit_display]
      if (length(levels(units))>1) {
        message(sprintf("WARNING: multiple units: %s", paste(levels(units), collapse = "; ")))
        report_stratstat$units[i] <- paste(levels(units), collapse = "; ")
      } else {
        report_stratstat$units[i] <- levels(units)[1]
      }
      report_stratstat$valcount[i] <- length(values)
      report_stratstat$sex[i] <- g
      report_stratstat$agegroup[i] <- a
      report_stratstat$mean[i] <- mean(values, na.rm=T)
      report_stratstat$median[i] <- median(values, na.rm=T)
      report_stratstat$sd[i] <- sd(values, na.rm=T)
    }
  }
  #write_delim(report_stratstat, sprintf("oracle_hf_data/stratified_stats_%s_report.tsv", lpm), "\t")
  report_stratstat %>%
    kbl(caption = sprintf("%s: Normal-physiological healthy patient cohort, stratified statistics", lpm), booktabs=T) %>%
    kable_styling(c("striped", "hover"), full_width = F, position="left")
}
```

